Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
tidying up to graphics section

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This section
discusses how these principles are applied, and proceeds in three parts. Section
\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
missing values that tracks missings throughout the analysis and allows for
special missing values. Section \@ref(graphics) covers visualizations to
understand missingness. Section \@ref(num-sum) discusses data summaries, and
Section \@ref(verbs) covers the common "verbs" used to solve problems
when working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.

@@

To explore missing data there needs to be a robust representation of missing
values that integrate into analysis pipelines. The R matrix, where 0 and 1
indicate not missing and missing provides a template, as it is very common in
missing data literature.  This matrix has been used to explore missing values in
the interactive graphics library XGobi, where it was called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness, where 1s represent missing values, and 0
represents complete values. The values 0 and 1 can
be confusing representations of missing values - does
0 indicate a the absence of observation, or the presence of a missing value? The
shadow matrix was specially linked to the data, but could not be referred to
directly, and was instead linked with special window views. This means
the data structure was abstracted from the user to facilitate data exploration.
This made sense at the time with the state of the art for graphical analysis,
but the data structure needs to be clearly presented to the user so they
can use it more flexibly in other data analyses.

@@

Four features are added to the shadow matrix to facilitate analysis:

@@

1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization and summaries.

@@

@@

## Missing value labels {#missing-labels}

The $R$ matrix is a very common representation of missing values, and can be easily created in the "R" programming language with \code{is.na(data)}. The values 0 and 1 (or equally, FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent - does 0 indicate a the absence of observation, or the presence of a missing value? This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).

## Special missings {#special-missings}

Values in the shadow matrix are "factor" type values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing. Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out". The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example. Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar}, shown in Section \@ref(verbs-recode). Special missing values do exist in other statistical programming languages, although are limited to 26 or 52 special types of missingness.

knitr::kable(fig_shadow_code,
             caption = "Example data of temperature, and its observed value, shadow representation, and underlying value. The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2. This captures additional information about the data otherwise difficult to record.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

The _nabular_ data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present. Firstly, it can be confusing to users whether TRUE or FALSE means missing or present; `NA` and `!NA` are clearer. Secondly, special missing values cannot be added fluently during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

@@

Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`. The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.

# Missing data operations {#verbs}

Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag. Missing values can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with an `NA` value. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**. These verbs are described in this section.

## scan: Searching for common missing value labels  {#verbs-search}

There are many common values for "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, all occurrences of -99 can be searched for with `miss_scan_count(data, search = -99)`. This returns a table of the occurrences of that value in each variable.

@@

A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers. These contain values like "n/a", "na ", and ".", and -9, -99. These can be used with `miss_scan_count` to help identify common missing values.

@@

Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values. This can be done with `replace_with_na()`. This is useful when there is certainty of which values mean missing. For example "N/A", "N A", "Not Available", and "missing". 
A dataset could have the values -99 in the `x` column replaced with the following code:
`replace_with_na(dat_ms, replace = list(x = -99))`

For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`.

@@

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)).

@@

An example use of these features is in @Tierney2015,
where the proportion, number, or cluster of missings is used as the outcome in a
model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`.

@@

_Nabular_ data has the shadow matrix column-bound to the existing data. This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked. _Nabular_ data can be created with `nabular()` or `bind_shadow()`. (The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe). An example use of `nabular()` some example data set is shown below:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using the `recode_shadow()` function. This provides a fluent interface to recode, or flag, the shadow information in the shadow matrix as a special type of missing value. Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.  These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis. The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

@@

Similar to `simputation`, each `impute_` function returns the data with values imputed. However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`. Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`). The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame. One drawback to imputation functions in this form is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

@@

To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`. The missing values can then be referred to by their shadow variable, `_NA`. The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)). The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_knn`, and adding a `label_shadow`:

@@

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C). Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows. This shows mean and median values are similar, but the minimum and maximum values are different.

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange. Imputed values are similar, but slightly different to existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses easy to remember visualizations that compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-graphics)); univariate (Section \@ref(univariate-graphics)), bivariate (Section \@ref(bivariate-graphics)), and multivariate (Section \@ref(multivariate-graphics)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.

@@

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables and cases for the airquality data. (A) Missings in each variable and (B) in each case. There are missing values in Ozone and Solar.R, with Ozone having more missings, and not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}

>>>>

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This
section discusses how these principles, in three parts: tidy data structures for missing data \@ref(data-structure), common operations (verbs) of missing data \@ref(verbs), graphics \@ref(graphics), and data summaries \@ref(num-sum). The design and naming of these features is discussed throughout, as they are an important component to help the analyst compose them efficiently in data pipelines.

@@

To explore missing data there needs to be a robust representation of missing
values that integrate into analysis pipelines. This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis. The R matrix, where 0 and 1
indicate not missing and missing provides a useful template, as it is very common in
missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness, where 1s represent missing values, and 0
represents complete values. The values 0 and 1 can
be confusing representations of missing values - does
0 indicate a the absence of observation, or the presence of a missing value? The
shadow matrix was specially linked to the data, but could not be access
directly, instead linked with special window views. The data structure was abstracted from the user to facilitate data exploration, which was consistent with the state of the art. However, the data structure needs to be clearly presented to the user so they
can apply it more flexibly in data analysis.

@@

We propose a new form for tidy representation of missing data. It is simple, and flexible, building off of past research. Four features are added to the shadow matrix to facilitate analysis of missing data:

@@

1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization, summaries, and tracking imputed values.

@@

### Missing value labels {#missing-labels}

The $R$ matrix is a very common representation of missing values. It can be created in the "R" language with \code{is.na(data)}. The values 0 and 1 (or FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent: does 0 indicate a the absence (0) of observation, or the presence (1) of a missing value? This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing. Values in the shadow matrix can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

### Special missings {#special-missings}

Extending missing value labels to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out", and even imputations (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`. The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.

knitr::kable(fig_shadow_code,
             caption = "Example data of temperature, its shadow representation, and underlying NA value. The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2. This captures additional information about the data otherwise difficult to record.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created with `nabular(data)` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

The _nabular_ data structure for missing data provides three key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present. Firstly, missing value labels `NA` and `!NA` are clearer. Secondly, special missing values cannot be added fluently during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

@@

## Missing data operations {#verbs}

Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag. Missing values can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with an `NA` value. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

### scan: Searching for common missing value labels  {#verbs-search}

There are many common values for "missing". For example, "N/A", "MISSING", `-99`. Before replacing these values with missing values (`NA`), it is useful to _scan_ the data to understand the magnitude of the problem.  `miss_scan_count()` finds where specified values occur in data. For example, a scan of all occurrences of -99 is done with `miss_scan_count(data, search = -99)`, returning a table of occurrences of that value for each variable. `common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters. They contain values like "n/a", "na", and -99, and are designed to be used with `miss_scan_count` to identify common missing values.

@@

@@

Once the magnitude of possible missing values has been assessed, these values need to be replaced with missing values. This can be done with `replace_with_na()`.  A dataset could have the values -99 in the `x` column replaced with the following code:
`replace_with_na(dat_ms, replace = list(x = -99))` For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).

@@

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. `naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`.

An example use of these features is in @Tierney2015, where the proportion of
missings is used as the outcome in a model, to identify variables and values
important in predicting missingness structures.

@@

@@

_Nabular_ data has the shadow matrix column-bound to the existing data. This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked. _Nabular_ data can be created with `nabular()`. An example use of `nabular()` some example data set is shown below:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using `recode_shadow()`. Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. Instead, _nabular_ data is designed to work with these. `naniar` does provide a few imputation methods to facilitate exploration and visualizations: `impute_mean`, `impute_median`, and `impute_below`. While useful to explore structure in missingness, they are not recommended for use in analysis. `impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

@@

Similar to `simputation`, each `impute_` function returns the data with values imputed. However, `naniar` does not use a formula syntax, instead each function has "scoped variants" `_all`, `_at` and `_if` as in \@ref(verbs-replace-with). `impute_` functions with no scoped variant, (`impute_mean`), will work on a single vector, but not a data.frame. One challenge with this approach is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

@@

To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`. The missing values can then be referred to by their shadow variable, `_NA`. The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)). The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_lm`, and adding a `label_shadow`:

@@

Missing values can be mapped to a graphical element in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C). Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows (shown below). This shows mean and median values are similar, but the minimum and maximum values are different.

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model. Imputed values are colored green, and data values orange. Imputed values are similar, but slightly trended to the mean of existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses four different areas of visualization of missing data: overviews (\@ref(overview-graphics)), univariate (\@ref(univariate-graphics)), bivariate (\@ref(bivariate-graphics)), and multivariate (\@ref(multivariate-graphics)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.

@@

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables (A) and cases (B) for the airquality data. There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}
